---
title: "Monte Carlo Integration with Adaptive Partitioning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Monte Carlo Integration with Adaptive Partitioning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
What is Monte Carlo Integration?
Monte Carlo integration estimates integrals by randomly sampling points and averaging function values. While simple and robust, it can be inefficient for functions with rapid variations or localized features. This package implements adaptive partitioning that automatically divides the integration domain into regions, focusing sampling effort where the function changes most rapidly.
Key Functions
montecarlo_integrate()
The main integration function with these parameters:

f: Function to integrate (must accept appropriate number of arguments)
lower: Lower bounds (scalar for 1D, vector for multi-dimensional)
upper: Upper bounds (scalar for 1D, vector for multi-dimensional)
n_samples: Total number of samples to use
Partition: Logical - use adaptive partitioning? (default: FALSE)
dim: Dimension of integration (default: 1)
step: Interval for recording convergence estimates (default: 100)

mc_print()
Print the final Monte Carlo estimate:

x: An mc_result object, list with $estimates, or numeric vector

mc_plot()
Visualize convergence of Monte Carlo estimates:

...: Named mc_result objects to compare (e.g., method1 = result1, method2 = result2)
step: Sample interval for x-axis scaling (default: 1)
true_value: Optional true integral value to show as reference line



```{r setup}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(StratMonteCarlo)
set.seed(123)
```


```{r basic-example}
## One dimensional example - simple polynomial
f1 <- function(x) x^2  # ∫[0,1] x² dx = 1/3
n_samples <- 50000
mc_f1 <- montecarlo_integrate(f1, 0, 1, n_samples, Partition = FALSE)
mc_f1_part <- montecarlo_integrate(f1, 0, 1, n_samples, Partition = TRUE)

mc_print(mc_f1)      # Should be ≈ 0.333
mc_print(mc_f1_part) # Should be ≈ 0.333
mc_plot(Standard = mc_f1, Partitioned = mc_f1_part, true_value = 1/3)
```

```{r oscillatory-example}
## Oscillatory function - where partitioning really helps
f_osc <- function(x) sin(10*x)  # ∫[0,2π] sin(10x) dx ≈ 0
mc_osc <- montecarlo_integrate(f_osc, 0, 2*pi, n_samples, Partition = FALSE)
mc_osc_part <- montecarlo_integrate(f_osc, 0, 2*pi, n_samples, Partition = TRUE)

cat("Oscillatory function (true value ≈ 0):\n")
mc_print(mc_osc)      # Often inaccurate
mc_print(mc_osc_part) # Much more accurate with partitioning
mc_plot(Standard = mc_osc, Partitioned = mc_osc_part, true_value = 0)
```

```{r 2d-simple}
## Two dimensional example - simple function
f2d <- function(x, y) x * y  # ∫[0,1]×[0,1] xy dx dy = 1/4
n_samples <- 40000
mc_2d <- montecarlo_integrate(f2d, c(0,0), c(1,1), n_samples, Partition = FALSE, dim = 2)
mc_2d_part <- montecarlo_integrate(f2d, c(0,0), c(1,1), n_samples, Partition = TRUE, dim = 2)

cat("2D integration (expected = 0.25):\n")
mc_print(mc_2d)
mc_print(mc_2d_part)
mc_plot(Standard = mc_2d, Partitioned = mc_2d_part, true_value = 0.25)
```

```{r 2d-peak}
## Two dimensional example - sharp peak (partitioning helps)
f_peak <- function(x, y) exp(-((x-0.5)^2 + (y-0.5)^2)/0.01)  # Sharp peak at (0.5, 0.5)
n_samples <- 20000
mc_peak <- montecarlo_integrate(f_peak, c(0,0), c(1,1), n_samples, Partition = FALSE, dim = 2)
mc_peak_part <- montecarlo_integrate(f_peak, c(0,0), c(1,1), n_samples, Partition = TRUE, dim = 2)

cat("Sharp peak function:\n")
mc_print(mc_peak)      # Less accurate
mc_print(mc_peak_part) # More accurate with partitioning
mc_plot(Standard = mc_peak, Partitioned = mc_peak_part)
```

```{r 3d-example}
## Three dimensional example
f3d <- function(x, y, z) x * y * z  # ∫[0,1]³ xyz dx dy dz = 1/8
n_samples <- 60000
mc_3d <- montecarlo_integrate(f3d, c(0,0,0), c(1,1,1), n_samples, Partition = FALSE, dim = 3)
mc_3d_part <- montecarlo_integrate(f3d, c(0,0,0), c(1,1,1), n_samples, Partition = TRUE, dim = 3)

cat("3D integration (expected = 0.125):\n")
mc_print(mc_3d)
mc_print(mc_3d_part)
```
